Estimating Daily Bicycle Counts in Seattle from Seasonal and Weather Factors

### 2014-12-08 article/article.tex
Finish draft results sections

\title{Estimating Bicycle Counts in Seattle from Seasonal and Weather Factors}
\author{Peter Schmiedeskamp, Weiran Zhao}
%\date{8 October 2014} % Delete this line to display the current date

@@

\subsection*{Model estimation}
% Model choice
A natural model choice for count data is the Poisson model, however we
believed that our count data were overdispersed. Overdispersion, or
contagion between events, violates the mean-variance equivalence
assumption of the Poisson model. While overdispersion would not impact
our parameter estimates, it would result in overly optimistic margins
of error. In order to account for and estimate the amount of
overdispersion present in our data, we chose a negative binomial model
type. Because there were no days observed with zero bicycle counts, we
did not need to resort to zero-inflated models as is often necessary
when dealing with count data.

@@

As seen in figure \ref{fg:seasonality}, we see a substantial increase
in bicycle volumes when the University of Washington is in session.
With all other factors held constant we see that, on days when the
university is in session, there is an average of approximately 367
additional bicycles observed. Similarly, we see a roughly linear
increase in bicycles with increased day length.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c4}
  \caption{Effect of daylight hours and University of Washington
    in-session status on bicycle counts, with shaded 95\% confidence
    regions.}
  \label{fg:seasonality}
\end{figure}

\subsection*{Weather}

\begin{figure}[h!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c2}
  \caption{Effect of precipitation on counts, with shaded 95\%
    confidence region.}
  \label{fg:precipitation}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c1}
  \caption{Effect of temperature on counts, with shaded 95\%
    confidence region.}
  \label{fg:temperature}
\end{figure}

\subsection*{General trend in bicycle counts}

\begin{figure}[h!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c5}
  \caption{General trend in bicycling counts, all other factors held
    constant, with 95\% confidence region.}
  \label{fg:trend}
\end{figure}

\subsection*{Day of week variation}

\begin{figure}[h!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c3}
  \caption{Variation in counts throughout the week, with 95\% confidence bars.}
  \label{fg:dayofweek}
\end{figure}

>>>>

% Float barriers
\usepackage[section]{placeins}

\title{Estimating Daily Bicycle Counts in Seattle from Seasonal and Weather Factors}
\author{Peter Schmiedeskamp, Weiran Zhao}
%\date{8 October 2014} % Delete this line to display the current date

@@

\subsection*{Model estimation and goodness of fit}
% Model choice
A natural model choice for count data is the Poisson model, however we
believed that our count data were overdispersed. Overdispersion, or
contagion between events, violates the mean-variance equivalence
assumption of the Poisson model. While overdispersion would not impact
our parameter estimates, it would result in overly optimistic margins
of error. In order to account for and estimate the amount of
overdispersion present in our data, we chose a negative binomial model
type. Because there were no days observed with zero bicycle counts, we
did not need to resort to zero-inflated models as is often necessary
when dealing with count data.

@@

Figure \ref{fg:seasonality} shows that we see a substantial increase
in bicycle volumes when the University of Washington is in session.
With all other factors held constant we see that, on days when the
university is in session, there is an average of approximately 367
additional bicycles observed. Similarly, we see a roughly linear
increase in bicycles with increased day length.

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c4}
  \caption{Effect of daylight hours and University of Washington
    in-session status on bicycle counts, with shaded 95\% confidence
    regions.}
  \label{fg:seasonality}
\end{figure}

\FloatBarrier
\subsection*{Weather}
As in the case of seasonality, our model represents weather using two
variables, in this case precipitation (measured as the maximum amount
of precipitation falling in any hour of that day), and maximum
temperature (measured daily, in Fahrenheit).

The effect of precipitation is shown in figure \ref{fg:precipitation}.
From this, we can see a clear inverse relationship between
precipitation and bicycle counts. The rate of decrease in bicycles
appears to begin somewhat steeply, and then begins to slow slightly at
higher amounts of precipitation. This suggests that people are
generally more sensitive to the presence of precipitation than the
intensity.

Temperature, in contrast to precipitation, has a clearly positive
association with increased numbers of bicyclists. We were somewhat
surprised to not see a leveling off in counts at very high
temperatures---likely due to Seattle's moderate summer climate, and the
lack of extremely warm days in our dataset.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c2}
  \caption{Effect of precipitation on counts, with shaded 95\%
    confidence region.}
  \label{fg:precipitation}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c1}
  \caption{Effect of temperature on counts, with shaded 95\%
    confidence region.}
  \label{fg:temperature}
\end{figure}

\FloatBarrier
\subsection*{General trend in bicycle counts}
Our results (shown in figure \ref{fg:trend}) confirm the presence of a
general trend toward increased numbers of bicycles at this location.
With all else held constant, we would expect to see roughly 300 more
bicycles on days at the end of our study period than at the beginning.
This is consistent with figures reported elsewhere that suggest that
bicycling is increasing in a number of cities including
Seattle \parencite{League-of-American-Bicyclists:aa}.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c5}
  \caption{General trend in bicycling counts, all other factors held
    constant, with 95\% confidence region.}
  \label{fg:trend}
\end{figure}

\FloatBarrier
\subsection*{Day of week variation}
As discussed previously, day of the week was included due to the
weekly variation in bicycle counts apparent in timeseries plots.
Figure \ref{fg:dayofweek} shows several interesting aspects of these
results. First, we see much higher numbers of bicyclists on weekdays
than on weekends. This strongly suggests that the majority of the
bicycle traffic at this location is for commuter purposes.

Comparing between weekday results, we see that most days are roughly
the same, with the some drop-off toward Thursday and Friday. This
decrease toward the end of the work week might be attributable to
individuals either working non-traditional schedules or perhaps people
adjusting their travel mode choice in order to accommodate social
engagements.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.5\textwidth]{../fig/m1c3}
  \caption{Variation in counts throughout the week, with 95\% confidence bars.}
  \label{fg:dayofweek}
\end{figure}
